---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

 [![Travis build status](https://travis-ci.com/baruuum/relabelKL.svg?token=k7R3D8yhYkrGz6yc4eQf&branch=master)](https://Travis-ci.com/baruuum/relabelKL)
 [![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/baruuum/relabelKL?branch=master&svg=true)](https://ci.appveyor.com/project/baruuum/relabelKL)
 [![Coverage status](https://codecov.io/gh/baruuum/relabelKL/branch/master/graph/badge.svg)](https://codecov.io/github/baruuum/relabelKL?branch=master)
 [![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)

## C++ implementation of the relabeling method proposed by Stephens(2000)

While the relabeling algorithm proposed in Stephens (2000) has been shown to perform well in dealing with the label switching phenomenon in Bayesian finite mixture models and has the attractive feature of requiring no additional information beyond the posterior samples, it has been pointed out that it is computationally expensive. Currently, the `label.switching::stephens` function provides the only `R` implementation of Stephens' method. Yet, the function is rather slow, which makes it impractical to use on MCMC output of moderate size.

The `relabelKL` package is a `C++` implementation of Stephen's relabeling algorithm. It is implemented using the [Armadillo library](http://arma.sourceforge.net/) and sourced via the `RcppArmadillo` package. If available, the functions provided will use OpenMP to run parts of the code in parallel. The package provides also a set of helper functions to facilitate the integration with `stanfit` objects produced via the `rstan` package.

A comparison between the performance of the `label.switching::stephens` function and the `relabelKL::relabelMCMC` function is provided at the end of this document.


## Installing the package

Using the `devtools` package, the developmental version of `relabelKL` can be installed from github:

``` r
devtools::install_github("baruuum/relabelKL")
library(relabelKL)
```

Mac OS users might experience compiling issues due to the OpenMP dependency. To set up the machine and install the necessary compiling tools, follow the instructions in [this page](https://thecoatlessprofessor.com/programming/cpp/r-compiler-tools-for-rcpp-on-macos/).


## Using the Package

Relabeling MCMC output is done by the `relabelMCMC` function. The function takes as input an array of dimensions `S`\*`N`\*`K`, where `S` is the number of posterior draws, `N` the number of units/individuals, `K` the latent classes/categories/types to which they belong, and the entries are either probabilities or log-probabilities of unit `i = 1,2,...,N` belonging to one of the `k=1,2,..,K` classes/types. 

Three options might be specified: 

1. `log_p`: whether the entries are log-probabilities or probabilities (defaults to `log_p = TRUE`)
2. `renormalize`: if `TRUE` renormalizes the rows of each slice of `x` to sum to one
3. `maxit`: the number of maximum iterations (defaults to `maxit = 100`)
4. `nthreads`: number of threads to use in calculation (if OPENMP is enabled)
5. `verbose`: whether intermediate results should be printed (defaults to `verbose = TRUE`); and

Calling `relabelMCMC` on the array `x`
``` r
res = relabelMCMC(x)
```
returns a `list` of four elements: 

1. `permuted`: the same object as `x` but with the columns of each posterior draw relabeled
2. `perms`: a `S \times K` matrix which contains, for each draw, the optimal permutation. That is, the `s`th row of `perms` shows how the `s`th posterior draw was permuted, so that applying the permutation in `perms` to the initial array, `x`, will result in the relabeled array
3. `iterations`: the number of iterations for which the algorithm was run
4. `status`: which is `0` if the algorithm has successfully converged and `1` otherwise


## Relabeling other parameters of the model

There are often other parameters in the model that depend on the labeling of the latent classes / extreme types. The `permuteMCMC` function can be used in this situation. For an three-dimensional array, `y`, where the last dimension correspond to `S` posterior draws, the `permuteMCMC` can be called based on the results from the `relabelMCMC` output (or for any other relabeling algorithm that returns the permutation mapping). Calling, for example,
```r
y_relabeled = permuteMCMC(y, perms = res$perms, what = "dimension to permute")
```
will permute the either the rows, columns, or both the rows and columns of each of the `S` sub-arrays of `y` based on `res$perms`. 

1. Three strings might be passed to the `what` option: specifying `"rows`" will permute the rows of each `s=1,2,...,S` draw, specifying `"cols"` will permute the columns, and `"both"` will permute both the rows *and* columns.
1. It is important to notice that the `permuteMCMC` function assumes that the first index of a three dimensional array represents the posterior draws, so that entering an array of dimensions, say,  `N`\*`B`\*`S` will lead to undefined behavior. 
2. When a two-dimensional matrix, instead of a three-dimensional array, is passed to `permuteMCMC`, it is assumed that it has dimensions `S`\*`K` and, the function will always permute the columns. 

## Relabeling when true assignment probabilities are known

If the "true" labels of a stochastic blockmodel or latent class model are known in advance, assigning each individual to their true class is straightforward. Yet, there are situations in which we want to make the assignment probabilities of each posterior draw as close as possible to a set of known/true probabilities. These situations arise, for example, when MAP or MLE estimates of the membership vectors of mixed membership models are calculated and when we want to use them as a pivots to relable the MCMC samples. In these cases, the `relabelTRUE` function might be used as follows:

```r
rel_true = relabelTRUE(x = x, x_true = x_true, verbose = F, log_p = F)
```

where `x_true` is a `N`\*`K` matrix of "true" assignment probabilities / mixed membership vectors. This function will try to relabel each posterior draw in `x` as close as possible to `x_true` in terms of KL-distances. When the `log_p` option is set to `TRUE`, *both* `x` and `x_true` should be entered on the log-scale. Otherwise, the function will throw an error.

## Functions to use with `rstan` objects

The package comes with one dataset called `mmsbm`, which is a `stanfit` object obtained from running a mixed membership stochastic blockmodel on simulated data. The model has two parameters: `pi`, a matrix the mixed membership vectors, where each row indicates the probability of individual `i = 1,2,..,N` enacting type `k = 1, 2, ..., K` in the interaction process, and `theta` the so-called image matrix, which is of dimensions `K`\*`K` and reflects the association tendencies between the `K` pure types. 

The data can be loaded as
```{r}
library(relabelKL)
data(mmsbm)
```
The `relabelKL` package provides a wrapper function to extract posterior samples from a `stanfit` object and rearrange the parameter draws so that it can be directly passed to the `relabelMCMC` function. 

Calling
```{r}
pi_arr = extract_n_combine(mmsbm, par = "pi")
```
will combine post-warmup draws across all MCMC chains into the first dimension of a three-dimensional array, where the other dimensions are identical to the dimensions specified in the `Stan` program. For example, if `pi` was specified as a `matrix[L,M]` object, the first dimension of `pi_arr` will be of order `S*J`, where `S` is as before the number of post-warmup draws and `J` the number of MCMC chains, and the two last dimensions of the extracted object will be `L` and `M`. 

For the `mmsbm` dataset, calling `extract_n_combine` results in an array of dimensions
```{r}
dim(pi_arr)
```

This object, then, can be passed to `relabelMCMC` or `permuteMCMC` as follows:

```{r}
rel = relabelMCMC(pi_arr, maxit = 50L, verbose = TRUE, log_p = FALSE)
```

The `rel$permuted` will contain the relabeled `pi` array, and we can apply the same permutations on other parameters by using the `rel$perms` object, which contains the permutations applied to the original MCMC output:
```{r}
# get permuted labels
rel_pi_arr = rel$permuted

# extract image matrix from stanfit object
theta_arr = extract_n_combine(mmsbm, par = "theta")

# apply same permutation to theta
rel_theta_arr = permuteMCMC(theta_arr, rel$perms, "both")
```

Lastly, to monitor the convergence of the relabeled posterior draws, we have to transform the relabeled arrays into a form that can be passed to the `rstan::monitor` function. This can be done by using the `to_stan_array` function as follows:
```{r}
# monitor convergence of original output
rstan::monitor(to_stan_array(theta_arr))

# same statistics for the relabeled output
rstan::monitor(to_stan_array(rel_theta_arr))
```
Similarly, `traceplot` might be inspected by using
```{r}
array_traceplot(to_stan_array(rel_theta_arr), par = "theta")
```
The call will produce a `ggplot` object containing the traceplot of the single element `pi[1,2]` of the `pi` matrix; the second call will produce a traceplot of all elements in the `theta` parameter.

## Comparison to `label.switching::stephens`

A simple comparison with the `stephens` function of the `label.switching` package might offer some insights into the efficiency gains. We first draw some random vectors that sum to one using the Dirichlet distribution.

```{r}
# set seed and dimensions
set.seed(123)
x = extract_n_combine(mmsbm, "pi")
```

Now, rigorous benchmarking will be super time-consuming; but as the difference in speed are quite large, we might use the `Sys.time()` function to get a sense of the efficiency gain. First, we use the `stephens` function of the `label.switching` package:

```{r}
# fit label.switching::stephens
start_t = Sys.time()
res1 = label.switching::stephens(x)
end_t = Sys.time()
print(end_t - start_t)
```

Next, we use the `relabelMCMC` function:

```{r}
# fit relabelKL::relabelMCMC
start_t = Sys.time()
res2 = relabelKL::relabelMCMC(x, maxit = 100L, verbose = FALSE, log_p = FALSE)
end_t = Sys.time()
print(end_t - start_t)
```

and with multiple threads

```{r}
# fit relabelKL::relabelMCMC
start_t = Sys.time()
res3 = relabelKL::relabelMCMC(x, maxit = 100L, verbose = FALSE, nthreads = 2L, log_p = FALSE)
end_t = Sys.time()
print(end_t - start_t)
```

Lastly, we check can check that the results are indentical:
```{r}
sum(res1$permutations != res2$perms)
sum(res1$permutations != res2$perms)
```


## References

Stephens, M. 2000. "Dealing with label Switching in mixture models," *Journal of the Royal Statistical Society Series B*, 62, 795-809.
